---
title: "Easy-to-use, intuitive, and efficient LC-MS data plotting with lcmsPlot"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2  
    number_sections: true  
    toc_float: true
vignette: >
  %\VignetteIndexEntry{basic_plotting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_knit$set(
  root.dir = dirname(getwd())
)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(openxlsx)
  library(xcms)
  library(BiocStyle)
  library(lcmsPlot)
})
```

# Introduction

LC-MS (or generally X-MS) data visualisation is important when inspecting the quality of the data
as well as during the publication process.
Plots are a quick way to show acquisition anomalies such as retention time shifts,
internal standards inconsistencies, gradual signal drops, and so on.

# Current landscape

The current LC-MS data visualisation landscape is mainly covered by `xcms`'s
base R plotting routines.

Outside `xcms`, many plotting functionalities are tightly coupled specific applications, such as:

- [LCMSQA](https://cran.r-project.org/web/packages/LCMSQA/index.html)
- [MetaboAnalystR](https://www.metaboanalyst.ca/docs/RTutorial.xhtml)
- [CAMERA](https://www.bioconductor.org/packages/release/bioc/html/CAMERA.html)

# Objectives

`lcmsPlot` aims to be a generic, flexible, and interoperable solution for plotting LC-MS data, such
as chromatograms, mass traces, mass spectra, and more.

`lcmsPlot` focuses on the following objectives:

- Generic and flexible plotting capabilities.
- Easy-to-use programming interface using a building-blocks approach such as [ggplot](https://ggplot2.tidyverse.org/).
- Interoperability: it can be seamlessly used with `xcms` or raw data (e.g., from mzML).
- Efficiency: better performance when scaling, compared to current solutions.

# Installation

`lcmsPlot` is a Bioconductor package and to install it we have to use [BiocManager](https://cran.r-project.org/web/packages/BiocManager/index.html).

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("lcmsPlot")
```

# Load data

In this vignette, we will use a set of plasma samples from a long-term stability study from the [Phenome Centre Birmingham](https://www.birmingham.ac.uk/research/centres-institutes/phenome-centre-birmingham).

The data was processed using [xcms](https://www.bioconductor.org/packages/release/bioc/html/xcms.html) and the
`XCMSnExp` object saved to `inst/extdata/xdata.rds`.

```{r}
xdata <- readRDS(system.file("extdata", "xdata.rds", package = "lcmsPlot"))
```

# Samples summary

## Plot base peak chromatogram (BPC)

Base peak chromatograms (BPCs) are useful in LC-MS because they simplify complex data and help quickly identify the most intense signals.

In the example below we plot the BPC for five samples:

```{r}
samples_metadata <- read.xlsx(system.file("extdata", "metadata.xlsx", package = "lcmsPlot"))

raw_files <- c(
  system.file("extdata", "data", "Human_Plasma_HILP_Ch1_Col1_scIC_062.mzML", package = "lcmsPlot"),
  system.file("extdata", "data", "Human_Plasma_HILP_Ch1_Col1_scIC_064.mzML", package = "lcmsPlot"),
  system.file("extdata", "data", "Human_Plasma_HILP_Ch1_Col1_scIC_066.mzML", package = "lcmsPlot"),
  system.file("extdata", "data", "Human_Plasma_HILP_Ch1_Col1_scIC_068.mzML", package = "lcmsPlot"),
  system.file("extdata", "data", "Human_Plasma_HILP_Ch1_Col1_scIC_070.mzML", package = "lcmsPlot")
)

lcmsPlot(raw_files, metadata = samples_metadata) +
  chromatogram(ppm = 5, rt_tol = 10, aggregation_fun = "max") +
  facets(facets = "sample_id", ncol = 1)
```

We can also plot multiple chromatograms in the same plot:

```{r}
raw_files <- c(
  system.file("extdata", "data", "Human_Plasma_HILP_Ch1_Col1_scIC_062.mzML", package = "lcmsPlot"),
  system.file("extdata", "data", "Human_Plasma_HILP_Ch1_Col1_scIC_064.mzML", package = "lcmsPlot")
)

lcmsPlot(raw_files) +
  chromatogram(ppm = 5, rt_tol = 10, aggregation_fun = "max") +
  arrange(group_by = "sample_id") +
  labels(title = "Base peak chromatograms", legend = "Sample") +
  legend(position = "bottom")
```

## Plot total ion chromatogram (TIC)

Total ion chromatogram (TIC) plots the sum of all ion intensities detected at each time point, giving a complete picture of everything entering the detector; not just the most abundant ion.

```{r}
raw_files <- c(
  system.file("extdata", "data", "Human_Plasma_HILP_Ch1_Col1_scIC_062.mzML", package = "lcmsPlot"),
  system.file("extdata", "data", "Human_Plasma_HILP_Ch1_Col1_scIC_064.mzML", package = "lcmsPlot")
)

lcmsPlot(raw_files) +
  chromatogram(ppm = 5, rt_tol = 10, aggregation_fun = "sum") +
  facets(facets = "sample_id", ncol = 1)
```

## Intensity maps

2D intensity maps are a quick way to see patterns in the acquired data.

Below, we build intensity density maps from two samples:

```{r}
raw_files <- c(
  "inst/extdata/data/Human_Plasma_HILP_Ch1_Col1_scIC_062.mzML",
  "inst/extdata/data/Human_Plasma_HILP_Ch1_Col1_scIC_064.mzML"
)

lcmsPlot(raw_files) +
  intensity_map(mz_range = c(100, 105), rt_range = c(0, 200), density = TRUE) +
  facets(facets = "sample_id", ncol = 2)
```

# Plot chromatograms

Next, we will plot chromatograms for a feature detected by `xcms` in two samples.

```{r}
lcmsPlot(xdata, 'sample_id') +
  chromatogram(features = 'M70T581',
               sample_ids = c(
                 'Human_Plasma_HILP_Ch1_Col1_scIC_062',
                 'Human_Plasma_HILP_Ch1_Col1_scIC_064'),
               ppm = 5,
               rt_tol = 10,
               highlight_peaks = TRUE) +
  arrange(group_by = 'sample_id') +
  labels(title = "Feature M70T581", legend = "Sample") +
  legend(position = "bottom")
```

The above plot can be faceted on metadata factors; in this case we facet on `feature_id` and `sample_id`:

```{r}
lcmsPlot(xdata, 'sample_id') +
  chromatogram(features = c('M70T581', 'M85T587_1'),
               sample_ids = c('Human_Plasma_HILP_Ch1_Col1_scIC_062',
                              'Human_Plasma_HILP_Ch1_Col1_scIC_064'),
               ppm = 5,
               rt_tol = 10,
               highlight_peaks = TRUE,
               highlight_peaks_color = "#121212") +
  facets(facets = c('feature_id', 'sample_id')) +
  labels(legend = "Sample") +
  legend(position = "bottom")
```

On the other hand, gridded plots (`grid` function), arrange panels in a matrix defined by two variables, allowing structured comparisons across both dimensions.

```{r}
lcmsPlot(xdata, 'sample_id') +
  chromatogram(features = c('M70T581', 'M85T587_1'),
               sample_ids = c(
                 'Human_Plasma_HILP_Ch1_Col1_scIC_062',
                 'Human_Plasma_HILP_Ch1_Col1_scIC_064'),
               ppm = 5,
               rt_tol = 10,
               highlight_peaks = TRUE,
               highlight_peaks_color = '#f00') +
  grid(rows = 'feature_id', cols = 'sample_id')
```

In some cases, you may want to highlight specific regions of the extracted ion chromatogram, for example, to draw attention to a peak that hasn't been detected yet. This can be done using the `rt_line` functionality:

```{r}
lcmsPlot(xdata, 'sample_id') +
  chromatogram(features = 'M70T581',
                       sample_ids = c(
                         'Human_Plasma_HILP_Ch1_Col1_scIC_062',
                         'Human_Plasma_HILP_Ch1_Col1_scIC_064'),
                       ppm = 5,
                       rt_tol = 10) +
  facets(facets = 'sample_id') +
  rt_line(intercept = 580, line_type = 'solid', color = 'red') +
  rt_line(intercept = 575, line_type = 'dashed', color = 'blue') +
  rt_line(intercept = 587, line_type = 'dashed', color = 'blue')
```

# Plot chromatograms with mass traces

Until now, we've primarily focused on chromatograms to understand the abundance and retention behaviour of features across samples.
These chromatograms show when and how strongly a compound elutes, but they don't reveal whether the mass accuracy or consistency is stable across the peak or between samples; this is where the mass trace (bottom panel) becomes valuable.

```{r}
lcmsPlot(xdata, 'sample_id') +
  chromatogram(features = 'M70T581',
                       sample_ids = c(
                         'Human_Plasma_HILP_Ch1_Col1_scIC_062',
                         'Human_Plasma_HILP_Ch1_Col1_scIC_064'),
                       ppm = 5,
                       rt_tol = 10,
                       highlight_peaks = TRUE) +
  mass_trace() +
  arrange(group_by = 'sample_id') +
  labels(title = "Feature M70T581", legend = "Sample") +
  legend(position = "bottom")
```

As for plain chromatograms, we can facet a combination of chromatograms + mass traces on a factor, such as `feature_id`:

```{r}
lcmsPlot(xdata, 'sample_id') +
  chromatogram(features = c('M70T581', 'M85T587_1'),
                       sample_ids = c(
                         'Human_Plasma_HILP_Ch1_Col1_scIC_062',
                         'Human_Plasma_HILP_Ch1_Col1_scIC_064'),
                       ppm = 5,
                       rt_tol = 10,
                       highlight_peaks = TRUE) +
  mass_trace() +
  arrange(group_by = "sample_id") +
  facets(facets = 'feature_id', ncol = 2) +
  labels(legend = "Sample") +
  legend(position = "bottom")
```

Or multiple factors (`feature_id` and `sample_id`):

```{r}
lcmsPlot(xdata, 'sample_id') +
  chromatogram(features = c('M70T581', 'M85T587_1'),
               sample_ids = c(
                 'Human_Plasma_HILP_Ch1_Col1_scIC_062',
                 'Human_Plasma_HILP_Ch1_Col1_scIC_064'),
               ppm = 5,
               rt_tol = 10,
               highlight_peaks = TRUE,
               highlight_peaks_color = '#343434') +
  mass_trace() +
  facets(facets = c('feature_id', 'sample_id'))
```

# Plot chromatograms from raw files

Until now, we’ve been using an `XCMSnExp` object as our data source. However, `lcmsPlot` also supports extracting chromatograms directly from raw files, such as those in the `mzML` format.

```{r}
raw_files <- c(
  system.file("extdata", "data", "Human_Plasma_HILP_Ch1_Col1_scIC_062.mzML", package = "lcmsPlot"),
  system.file("extdata", "data", "Human_Plasma_HILP_Ch1_Col1_scIC_064.mzML", package = "lcmsPlot")
)

lcmsPlot(raw_files) +
  chromatogram(features = rbind(c(mz = 70.06507, rt = 580.642449), c(mz = 85.06472, rt = 587.131287)),
               ppm = 5,
               rt_tol = 10) +
  arrange(group_by = "sample_id") +
  facets(facets = "feature_id", ncol = 2) +
  legend(position = "bottom") +
  labels(legend = "Sample")
```

Metadata can be attached to the raw files and used for plotting as follows:

```{r}
raw_files <- c(
  system.file("extdata", "data", "Human_Plasma_HILP_Ch1_Col1_scIC_062.mzML", package = "lcmsPlot"),
  system.file("extdata", "data", "Human_Plasma_HILP_Ch1_Col1_scIC_064.mzML", package = "lcmsPlot")
)

metadata <- data.frame(
  sample_type = c("T1", "T2"),
  sample_id = c("S62", "S64")
)

lcmsPlot(raw_files, sample_id_column = "sample_id", metadata = metadata) +
  chromatogram(features = rbind(c(mz = 70.06507, rt = 580.642449), c(mz = 85.06472, rt = 587.131287)),
               ppm = 5,
               rt_tol = 10) +
  grid(rows = 'feature_id', cols = 'sample_type')
```

# Plot spectra

Now, we want to include spectra because they provide critical compositional insight into what ions make up the feature at a given point in time.

To do so, we can use the `spectra` function and define what mode of scan selection to use.

## Select closest scan to specified RT

To select the closest scan to a specified RT value you can use `mode = "closest"`:

```{r}
lcmsPlot(xdata, 'sample_id') +
  chromatogram(features = 'M70T581',
               sample_ids = 'Human_Plasma_HILP_Ch1_Col1_scIC_062',
               ppm = 5,
               rt_tol = 10,
               highlight_peaks = TRUE) +
  spectra(rt = 580, mode = "closest", ms_level = 1) +
  labels(title = "Feature M70T581", legend = "Sample") +
  legend(position = "bottom")
```

## Select scan closest to a detected peak apex

To select the closest scan to a detected peak apex you can use `mode = "closest_apex"`:

```{r, fig.height=3}
lcmsPlot(xdata, 'sample_id') +
  chromatogram(features = rbind(c(mz = 70.06507, rt = 580.642449), c(mz = 85.06472, rt = 587.131287)),
               sample_ids = 'Human_Plasma_HILP_Ch1_Col1_scIC_062',
               ppm = 5,
               rt_tol = 40,
               highlight_peaks = TRUE) +
  spectra(mode = "closest_apex", ms_level = 1) +
  facets(facets = "feature_id", ncol = 2) +
  labels(legend = "Sample") +
  legend(position = "bottom") +
  layout(design = "C\nS\nS")
```

## Select multiple scans across a detected peak

To select multiple scans across a detected peak, given an RT interval, you can use `mode = "across_peak"`:

```{r, fig.height=5, fig.width=3}
lcmsPlot(xdata, 'sample_id') +
  chromatogram(features = rbind(c(mz = 70.06507, rt = 580.642449)),
               sample_ids = 'Human_Plasma_HILP_Ch1_Col1_scIC_062',
               ppm = 5,
               rt_tol = 40,
               highlight_peaks = TRUE) +
  spectra(mode = "across_peak", ms_level = 1) +
  labels(legend = "Sample") +
  legend(position = "bottom") +
  layout(design = "C\nS\nS\nS\nS\nS\nS\nS")
```

## Plot standalone spectra

So far we have been plotting spectra alongside chromatograms.
Often, it is preferable to plot spectra on their own to allow for clearer analysis of spectral features:

```{r}
sample_ids = c(
  'Human_Plasma_HILP_Ch1_Col1_scIC_062',
  'Human_Plasma_HILP_Ch1_Col1_scIC_064'
)

lcmsPlot(xdata, 'sample_id') +
  spectra(sample_ids = sample_ids, rt = 580, mode = "closest", ms_level = 1) +
  legend(position = "bottom")
```

```{r}
sample_ids = c(
  'Human_Plasma_HILP_Ch1_Col1_scIC_062',
  'Human_Plasma_HILP_Ch1_Col1_scIC_064'
)

lcmsPlot(xdata, 'sample_id') +
  spectra(
    sample_ids = sample_ids,
    rt = 580,
    mode = "closest",
    ms_level = 1,
    spectral_match_db = dir(system.file("sciex", package = "msdata"), full.names = TRUE),
    match_target_index = 1) +
  legend(position = "bottom")
```

